### 2018-06-01 submission/manuscript.tex
updates from a final read through

\section{Error Introduced into 16S rRNA Gene Sequencing Results Varies
by High Fidelity DNA Polymerase
Used}\label{error-introduced-into-16s-rrna-gene-sequencing-results-varies-by-high-fidelity-dna-polymerase-used}

@@

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, chimera prevalence, and the correlation
between chimera prevalence and number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs varied by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variation in the number of OTUs detected by
the polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\textsuperscript{2} = 0.974) and Kappa having the worst
(R\textsuperscript{2} = 0.478).

@@

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors could reduce the difference between the two polymerases by
optimizing the annealing and extension steps within the PCR protocol
(Gohl et al., 2016). Yet, within this study there was no comparison made
between different high fidelity DNA polymerases. Due to this gap, it is
natural to extend this line of inquiry and test if biases in the number
of OTUs and chimeras also are dependent on the type of high fidelity DNA
polymerase. This study will investigate how high fidelity DNA
polymerases can bias observed bacterial community results derived from
16S rRNA gene sequencing. We will accomplish this by examining the
number of OTUs, error rate, number of sequences with an error, and
chimera prevalence at varying number of cycles in five different high
fidelity DNA polymerases

@@

\textbf{\emph{Human and Mock Samples:}} Fecal samples were obtained from
4 individuals who were part of the Enterics Research Investigational
Network (ERIN). The processing and storage of these samples were
previously published (Seekatz et al., 2016). Other than confirmation
that none of these individuals had a \emph{Clostridium difficle}
infection, clinical data and other types of meta data were not utilized
or accessed for this study. All samples were extracted using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR Protocol:}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
polymerases activation time was 2 minutes, unless a different activation
was specified by the manufacturer. The annealing and extension time for
Platinum and Accuprime followed a previously published protocol (Kozich
et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time were from a
previously published protocol (Gohl et al., 2016). For Phusion, the
company defined activation and annealing times were used while the
extension time followed the Accuprime and Platinum settings.

@@

\textbf{\emph{Sequence Processing:}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences (Schloss et al., 2009; Cole et al.,
2013; Rognes et al., 2016).

\textbf{\emph{Analysis Workflow:}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis. The
total number of OTUs was analyzed for both the fecal and mock community
samples. For fecal samples, cycle dependent affects on Bray-Curtis
indices were assessed for cycle group and within individual differences
from the previous cycle (e.g.~20 versus 25, 25 versus 30). Based on
these observations we analyzed potential reasons for these differences.
Analysis of the mock community of each polymerase for sequence error
rate, number of sequences with an error, base substitution, and numbers
of chimeras before and after chimera removal with VSEARCH was assessed.
Additionally, the correlation between the number of chimeras and the
number of OTUs was also assessed.

\textbf{\emph{Statistical Analysis:}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the fecal samples the data was
normalized to each individual by cycle number to account for the
biological variation between people. Bray-Curtis distance matrices were
generated using mothur after 100 sub-samplings at 1000, 5000, 10000, and
15000 sequence depth. The distance matrix data was analyzed using
PERMANOVA with the vegan package (v 2.4.5) (Oksanen et al., 2017) and
Kruskal-Wallis tests within R. For both error and chimera analysis,
samples were tested using Kruskal-Wallis with a Dunns post-hoc test.
Where applicable correction for multiple comparison utilized the
Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).

\textbf{\emph{Reproducible Methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found in the SRA at the following accession
number SRP132931.

@@

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used:}} Differences in the range of the number of OTUs detected for
fecal samples is dependent on the polymerase used (e.g.~Accuprime at 35
cycles = 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure
1{]}. Additionally, there is a trend for lower number of cycles (15-20)
to result in a reduced range in the number of OTUs versus higher number
of cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15
cycles = 10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
There is an overall difference in the number of OTUs detected within
fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
16.35, P-value = 9.7e-05) {[}Table S1{]}. Using a Tukey post-hoc test to
identify which polymerase groups were different from each other at 35
cycles, no difference was found (P-value \textgreater{} 0.05) {[}Table
S2{]}. The polymerase dependent difference in the range of the number of
OTUs also was observed in the mock community samples {[}Figure 2{]}. The
closest the polymerases came to the total of 8 OTUs created by the mock
reference 16S sequences was at 25 and 30 cycles {[}Figure 2{]}.
Regardless of if fecal or mock communities were used, the same
polymerases generated high and low number of OTUs and this was
consistent across the number of cycles used {[}Figure 1-2 \& Table
S1-S2{]}. In contrast to the results obtained with fecal samples,
differences between polymerases for the number of OTUs created were
observed as early as 20 cycles in the mock community (F-stat = 15.82,
P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc test, the
majority of differences for the number of OTUs detected in the mock
community is largely due to Kappa and Platinum differences versus the
other polymerases across the different number of cycles {[}Table S2{]}.
Based on these observations in fecal and mock communities, it is clear
that using different polymerases result in a different total number of
OTUs within a sample.

\textbf{\emph{The bacterial community is similar within polymerase and
varies by number of cycles:}} Within each respective 5-cycle increment
comparison there was no difference in Bray-Curtis index between the
polymerases (P-value \textgreater{} 0.05) {[}Figure 3{]}. Using
PERMANOVA to test for community differences based on any of the number
of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03. For fecal samples, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 at 15
cycles. There was an overall decrease in Bray-Curtis index by cycle
comparison group (e.g.~the 15 cycle versus 20 cycle group compared to
the 30 cycle to 35 cycle group) (P-value \textless{} 0.01) {[}Figure
3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
cycle versus 25 cycle comparison groups had a higher Bray-Curtis index
than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
The mock community has similar trends to the observations reported for
the fecal bacterial community but none were significant \textgreater{}
0.05) {[}Figure 3B{]}. Overall, these data suggest that the number of
cycles can change the bacterial community independent of the polymerase
used to generate the sequences.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used:}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 and Table S4{]}. The total
sequences with at least one error is also polymerase dependent with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S3 \& S4{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific base substitution.

@@

The authors would like to thank all the study participants in ERIN whose
samples were utilized. We would also like to thank Judy Opp and April
Cockburn for their effort in sequencing the samples as part of the
Microbiome Core Facility at the University of Michigan. Additional
thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
earlier drafts of the manuscript and providing helpful critiques. Salary
support for Marc A. Sze came from the Canadian Institute of Health
Research and NIH grant UL1TR002240. Salary support for Patrick D.
Schloss came from NIH grants P30DK034933 and 1R01CA215574.

>>>>

\section{Error Introduced into 16S rRNA Gene Sequencing Varies by High
Fidelity DNA Polymerase
Used}\label{error-introduced-into-16s-rrna-gene-sequencing-varies-by-high-fidelity-dna-polymerase-used}

@@

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase and number of cycle dependent differences in the Bray-Curtis
index, error rate, sequence error prevalence, chimera prevalence, and
the correlation between chimera prevalence and number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases (eg. at 35 cycles Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs varied by
HiFi DNA polymerase used at 35 cycles (P-value \textless{} 0.0001). When
analyzing mock community samples the variation in the number of OTUs
detected by the polymerases was observable as early as 20 cycles
(P-value = 0.002). There also was a large range in the number of OTUs
amplified by the polymerases at 35 cycles (Accuprime = 15 - 20 versus
Phusion = 14 - 73). Chimera prevalence in mock communities varied by
polymerase with differences being most notable at 35 cycles (Kappa =
5.71\% (median) versus Platinum = 26.62\%) and this variation persisted
after chimera removal using VSEARCH. We also observed positive
correlations between chimera prevalence and the number of OTUs with
Platinum having the highest (R\textsuperscript{2} = 0.974) and Kappa
having the worst (R\textsuperscript{2} = 0.478).

@@

A recent study found differences in the number of OTUS and chimeras
between normal and high fidelity DNA polymerases (Gohl et al., 2016).
The authors could reduce the difference between the two polymerases by
optimizing the annealing and extension steps within the PCR protocol
(Gohl et al., 2016). Yet, within this study there was no comparison made
between different high fidelity DNA polymerases. Due to this gap, it is
natural to extend this line of inquiry and test if biases in the number
of OTUs and chimeras also are dependent on the type of high fidelity DNA
polymerase. This study will investigate how high fidelity DNA
polymerases can bias observed bacterial community results derived from
16S rRNA gene sequencing. We will accomplish this by examining the
number of OTUs, Bray-Curtis index, error rate, number of sequences with
an error, and chimera prevalence at varying number of cycles in five
different high fidelity DNA polymerases

@@

\textbf{\emph{Human and mock samples:}} Fecal samples were obtained from
4 individuals who were part of the Enterics Research Investigational
Network (ERIN). The processing and storage of these samples were
previously published (Seekatz et al., 2016). Other than confirmation
that none of these individuals had a \emph{Clostridium difficle}
infection, clinical data and other types of meta data were not utilized
or accessed for this study. All samples were extracted using the
MOBIO\textsuperscript{TM} PowerMag Microbiome RNA/DNA extraction kit
(now Qiagen, MD, USA). The ZymoBIOMICS\textsuperscript{TM} Microbial
Community DNA Standard (Zymo, CA, USA) was used for mock communities and
was made up of \emph{Pseudomonas aeruginosa}, \emph{Escherichia coli},
\emph{Salmonella enterica}, \emph{Lactobacillus fermentum},
\emph{Enterococcus faecalis}, \emph{Staphylococcus aureus},
\emph{Listeria monocytogenes}, and \emph{Bacillus subtilis} at equal
genomic DNA abundance
(\url{http://www.zymoresearch.com/microbiomics/microbial-standards/zymobiomics-microbial-community-standards}).

\textbf{\emph{PCR protocol:}} The five different high fidelity DNA
polymerases (hereto referred to as polymerases) that were tested
included AccuPrime\textsuperscript{TM} (ThermoFisher, MA, USA), KAPA
HIFI (Roche, IN, USA), Phusion (ThermoFisher, MA, USA), Platinum
(ThermoFisher, MA, USA), and Q5 (New England Biolabs, MA, USA). The
polymerases activation time was 2 minutes, unless a different activation
was specified by the manufacturer. The annealing and extension time for
Platinum and Accuprime followed a previously published protocol (Kozich
et al., 2013)
(\url{https://github.com/SchlossLab/MiSeq_WetLab_SOP/blob/master/MiSeq_WetLab_SOP_v4.md}).
For Kappa and Q5, the annealing and extension time also were from a
previously published protocol (Gohl et al., 2016). For Phusion, the
company defined activation and annealing times were used while the
extension time followed the Accuprime and Platinum settings.

@@

\textbf{\emph{Sequence processing:}} The mothur software program was
used for all sequence processing steps (Schloss et al., 2009). The
protocol has been previously published (Kozich et al., 2013)
(\url{https://www.mothur.org/wiki/MiSeq_SOP}). Two major differences
from the published protocol were the use of VSEARCH instead of UCHIME
for chimera detection and the use of the OptiClust algorithm instead of
average neighbor for OTU generation at 97\% similarity (Edgar et al.,
2011; Rognes et al., 2016; Westcott \& Schloss, 2017). Sequence error
was determined using the `seq.error' command on mock samples to compare
back to the reference 16S sequences of \emph{P. aeruginosa}, \emph{E.
coli}, \emph{S. enterica}, \emph{L. fermentum}, \emph{E. faecalis},
\emph{S. aureus}, \emph{L. monocytogenes}, and \emph{B. subtilis}
(Schloss et al., 2009; Cole et al., 2013; Rognes et al., 2016).

\textbf{\emph{Analysis workflow:}} To adjust for unequal sequencing, all
samples were rarefied to 1000 sequences for downstream analysis. The
total number of OTUs was analyzed for both the fecal and mock community
samples. For fecal samples, cycle dependent affects on Bray-Curtis
indices were assessed for cycle group and within individual differences
from the previous cycle (e.g.~20 versus 25, 25 versus 30). Based on
these observations we analyzed potential reasons for these differences
using the mock community samples. Analysis of the mock community of each
polymerase for sequence error rate, number of sequences with an error,
base substitution, and numbers of chimeras before and after chimera
removal with VSEARCH was assessed. Additionally, the correlation between
the number of chimeras and the number of OTUs was also assessed.

\textbf{\emph{Statistical analysis:}} All analysis was done with the R
(v 3.4.4) software package (R Core Team, 2017). Data transformation and
graphing was completed using the tidyverse package (v 1.2.1) and colors
selected using the viridis package (v 0.4.1) (Garnier, 2017; Wickham,
2017). Differences in the total number of OTUs were analyzed using an
ANOVA with a tukey post-hoc test. For the comparison in fecal samples
the data was normalized to each individual by cycle number to account
for the biological variation between people. Bray-Curtis distance
matrices were generated using mothur. The distance matrix data was
analyzed using PERMANOVA with the vegan package (v 2.4.5) (Oksanen et
al., 2017) and Kruskal-Wallis tests within R. For both error and chimera
analysis, samples were tested using Kruskal-Wallis with a Dunns post-hoc
test. Where applicable, correction for multiple comparison utilized the
Benjamini-Hochberg method (Benjamini \& Hochberg, 1995).

\textbf{\emph{Reproducible methods:}} The code and analysis can be found
here \url{https://github.com/SchlossLab/Sze_PCRSeqEffects_XXXX_2017}.
The raw sequences can be found on the SRA at the following accession
number SRP132931.

@@

\textbf{\emph{The number of OTUs generated are dependent on polymerase
used:}} Differences in the range of the number of OTUs detected for
fecal samples is dependent on the polymerase used (e.g.~Accuprime at 35
cycles = 84 - 106 versus Phusion at 35 cycles = 84 - 136) {[}Figure
1{]}. Additionally, there is a trend for lower number of cycles (15-20)
to result in a reduced range in the number of OTUs versus higher number
of cycles (25, 30, and 35) for all polymerases (e.g.~Phusion at 15
cycles = 10 - 19 versus Phusion at 35 cycles = 84 - 136) {[}Figure 1{]}.
There is an overall difference in the number of OTUs detected within
fecal samples between polymerases at 35 cycles (F-stat \textgreater{}
16.35, P-value = 9.7e-05) {[}Table S1{]}, but a Tukey post-hoc test
failed to identify which specific polymerase groups were different
(P-value \textgreater{} 0.05) {[}Table S2{]}. The polymerase dependent
difference in the range of the number of OTUs also was observed in the
mock community samples {[}Figure 2{]}. The closest the polymerases came
to the total of 8 OTUs created by the mock reference 16S sequences was
at 25 and 30 cycles {[}Figure 2{]}. Regardless of if fecal or mock
communities were used, the same polymerases generated high and low
number of OTUs and this was consistent across the number of cycles used
{[}Figure 1-2 \& Table S1-S2{]}. In contrast to the results obtained
with fecal samples, differences between polymerases for the number of
OTUs created were observed as early as 20 cycles in the mock community
(F-stat = 15.82, P-value = 0.002) {[}Table S1{]}. Using a Tukey post-hoc
test, the majority of differences for the number of OTUs detected in the
mock community was largely due to Kappa and Platinum versus the other
polymerases across the number of cycles used {[}Table S2{]}. Based on
these observations in fecal and mock communities, it is clear that using
different polymerases result in a different total number of OTUs within
a sample.

\textbf{\emph{The bacterial communities generated by the polymerases
were similar but varied by number of cycles:}} Within each respective
5-cycle increment comparison there was no difference in Bray-Curtis
index between the polymerases (P-value \textgreater{} 0.05) {[}Figure
3{]}. Using PERMANOVA to test for community differences based on any of
the number of cycles within polymerases, only Phusion had cycle
dependent differences (P-value = 0.03). For fecal samples, Phusion was
one of two polymerases that had enough sequences to be rarefied to 1000
sequences at 15 cycles. There was an overall decrease in Bray-Curtis
index by cycle comparison group (e.g.~the 15 cycle versus 20 cycle group
compared to the 30 cycle versus 35 cycle group) (P-value \textless{}
0.01) {[}Figure 3A{]}. Using a Dunn's post-hoc test the 15 cycle versus
20 cycle and 20 cycle versus 25 cycle comparison groups had a higher
Bray-Curtis index than the 30 cycle versus 35 cycle comparison group
(P-value \textless{} 0.05). Similar trends were observed in the mock
community but none were significant (P-value \textgreater{} 0.05)
{[}Figure 3B{]}. Overall, these data suggest that the number of cycles
can change the bacterial community independent of the polymerase used to
generate the sequences.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used:}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 \& Table S4{]}. The total
sequences with at least one error was also polymerase dependent, with
the majority of differences being between Kappa or Accuprime and the
other polymerases {[}Table S3 \& S4{]}. These differences in error rates
were not due to polymerase dependent differences in base substitution
rate {[}Figure S1{]}. Collectively, the results suggest that sequence
error is dependent on polymerase, persists across the number of cycles
used, and are not due to any bias towards a specific base substitution.

@@

The authors would like to thank all the study participants in ERIN whose
samples were utilized. We also would like to thank Judy Opp and April
Cockburn for their effort in sequencing the samples as part of the
Microbiome Core Facility at the University of Michigan. Additional
thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
earlier drafts of the manuscript and providing helpful critiques. Salary
support for Marc A. Sze came from the Canadian Institute of Health
Research and NIH grant UL1TR002240. Salary support for Patrick D.
Schloss came from NIH grants P30DK034933 and 1R01CA215574.
